Demographic inference using the SFS with \moments and \demes

### 2025-08-26 chapter/paper.tex
final version of revision 1

\emph{Keywords:} Demographic inference, site frequency spectrum, diffusion
equation, moments equations

@@

Many evolutionary mechanisms are known to affect allele frequency dynamics and
the SFS. This presents a challenge for demographic inference, as the observed
SFS may be distorted by non-demographic processes. However, this also presents
an opportunity to learn about different evolutionary processes, if demographic
history can be controlled for. The SFS has been used to infer the distribution
of fitness effects of new mutations, primarily for nonsynonymous variation in
coding regions (e.g., \cite{eyre2006distribution, boyko2008assessing,
kim2017inference}); to scan for historical selective sweeps
(e.g., \cite{kim2002detecting, nielsen2005genomic}); and to understand how
selection on quantitative traits impacts the genetic architecture underlying
those traits (e.g., \cite{patel2024conditional, ragsdale2024archaic}). The SFS
has also been used to infer relative mutation rates and how they have changed
over time \cite{dewitt2021nonparametric}. Each of these analyses typically
requires partitioning the data in some way -- by functional annotation,
mutation context, local recombination rate, etc -- so that comparisons can be
made across SFS from different classes of mutations.

@@

\singlespacing
\begin{minted}[gobble=4, frame=single,]{yaml}
    description: split-with-migration model, loosely based on example 2
    generation_time: 29
    time_units: years
    demes:
    - name: ancestral
      epochs:
      - {start_size: 15000, end_time: 300000}
      - {start_size: 25000, end_time: 75000}
    - name: popA
      ancestors: [ancestral]
      epochs:
      - {start_size: 30000}
    - name: popB
      ancestors: [ancestral]
      epochs:
      - {start_size: 1200, end_size: 14000}
    migrations:
    - demes: [popA, popB]
      rate: 5e-5
\end{minted}
\doublespacing

In \moments, a \demes-specified demographic model can be directly used to
compute expectations for either the SFS or multi-population LD statistics
\cite{ragsdale2019models, ragsdale2020unbiased}. For the SFS, we simply need
to load the demographic model using \demes and then specify the number of
haploid samples to draw and from which populations.
\begin{minted}[gobble=4, frame=single]{python}
    import demes, moments
    g = demes.load('model.yaml')
    samples = {'popA': 60, 'popB': 60}
    fs = moments.Demes.SFS(g, samples=samples)
\end{minted}
Here, we have not specified a mutation rate. The default behavior sets $u=1$,
so that $\theta = 4N_eu=4N_e$, where $N_e$ is the ancestral population's
initial size. With a known mutation rate, the SFS will be properly scaled by
passing the mutation rate as a keyword argument: \texttt{moments.Demes.SFS(g,
samples=samples, u=u)}. This is equivalent to scaling the output SFS above by
multiplying by \texttt{u}, as \texttt{u*fs}.

@@

Below, we show two examples of multi-population demographic inference using the
joint SFS. The first example simulates data with known demographic parameters,
which we try to reinfer using both the original model and simpler
misspecified models. In the second example, we infer a relatively simple
Neanderthal-human history using data from two human populations and a
high-coverage ancient sample from the Neandertal lineage. For both of these
examples, we briefly describe data processing and inference setup, emphasizing
the high-level components of the analyses. For detailed information about each
example, including managing data, specifying models, performing inference,
computing confidence intervals and visualizing fits, we refer readers to the
accompanying GitHub repository
(\url{https://github.com/StatisticalPopulationGenomics-2ndEd/moments}), which
is maintained with up-to-date versioning.

@@

We used publicly available data to infer a demographic model for two
modern human populations (a northern European and western African population) 
and a Neandertal individual. As shown below, even with just a single sample 
from the Neandertal lineage, we are able to recover key parameters such as the
Neanderthal-human divergence time and the fraction of ancestry contributed to 
modern Europeans by Neandertals.

We obtained modern human genome sequences from a recent resequencing of the 
1000 Genomes Project cohort \cite{byrska2022high}, and used the high-coverage 
sequence of the Vindija Neandertal genome \cite{prufer2017high}. 
From the 1000 Genomes cohort we took 85 MSL (Mende from Sierre Leone) and 91 
GBR (British from England and Scotland) sequences as our modern human samples.  
We filtered out sites that lay outside the 1000 Genomes `strict' mask, 
were not genotyped in four high-coverage archaic genome sequences, 
lacked high-confidence ancestral state assignments, or fell within 10 kb of 
exonic or promoter regions. This left us with approximately 960 Mb of 
well-characterized and putatively neutrally-evolving sequence.
We estimated the genome-wide SFS using built-in \moments parsing functions.
Details concerning data processing (including our liftover of the Vindija 
sequence to a more recent genome build) can be found in the linked GitHub 
repository.

@@

We next stitched these one-population models together and refitted the resulting 
demographic model to get a two-population model that included continuous 
symmetric migration between MSL and GBR and an ancestral size expansion. 
Following this, we added a Neandertal branch and fitted the 
Neandertal-human divergence time, the Neandertal effective size, and the 
Neandertal-GBR pulse proportion, leaving all other parameters fixed. We also 
fitted the divergence time between the sampled Vindija Neandertal lineage and 
the Neandertal lineage which admixed with non-African humans. Finding 
that this parameter tended to converge to its lower bound (imposed by the 
estimated time of death of the Vindija Neandertal, ~55 kya), we fixed it to a 
value supported by literature (90 kya, \cite{prufer2017high}) in a second round 
of optimization. We also fixed the time of the Neandertal-GBR pulse to 50 kya 
(at the upper range of values inferrred by \cite{sumer2025earliest}). 
Fixing parameters for which good estimates are available in the literature is often 
reasonable in the first stages of model optimization, and we expected this 
value to be especially poorly constrained due to the small pulse proportion and 
Neandertal sample size. In the second round of optimization of the 
three-population model, we refitted all parameters aside from the two times 
mentioned above to refine the estimates.

After this refitting step, we were left with parameters in good agreement with 
those inferred by prior authors, with a Neandertal-human divergence time 
of 547 kya, MSL-GBR split time of 77 kya, and Neandertal-GBR pulse proportion 
of 2.7 percent. We observe some large, systematic residuals in our best-fit model 
(Figure~\ref{fig:humans}G-I), suggesting model misspecification that could be
improved through further exploration of model space.

We also fitted a nested model without the Neanderthal-GBR pulse, which had a 
substantially lower likelihood than the pulse model (log-likelihood [LL] 
$\approx -127,552$, against LL $\approx -25,103$ for the pulse model). We did 
not perform a formal hypothesis test; the null distribution of the 
likelihood-ratio test (LRT) statistic is overdispersed under composite 
likelihood, inflating the rate of false positives and preventing its use 
 in this setting. \moments does incorporate a function 
(\texttt{moments.Godambe.LRT\_adjust()}) which computes a multiplicative
adjustment to the LRT, intended to restore the expected null distribution, using 
first-order moment matching \cite{coffman2016computationally}. This method 
requires bootstrapping over the data. Otherwise, we could conduct a hypothesis 
test by fitting the nested and pulse models to a large number of bootstrap 
samples and calculating the rate of rejection across them-- which has the 
benefit of producing a distribution of (nonindependent) LRT statistics, but is
time-consuming and computationally expensive.

@@

\begin{figure}
    %\includegraphics{../example1/fig1.pdf}
    \caption{
        Inference using simulated data. (A) The true demographic model,
        including an expansion in the ancestral population and symmetric
        migration between the descendent populations. (B--D) The simulated
        data, showing both marginal and the joint SFS. (E--G) Residuals
        between the data SFS (in D) and the predicted SFS from the inferred
        demographic models in (H--J), resp.  The two misspecified models (H and
        I) provide worse fits to the data and both overestimate the split time
        due to disallowing any size change in the ancestral population.
    }
    \label{fig:im}
\end{figure}

\begin{figure}
    %\includegraphics{../example2/fig2.pdf}
    \caption{
        Inference using human and Neandertal data.
        (A) The best-fit model including the Vindija Neandertal and two human
        populations.
        (B, C) Marginal SFS fits between the model and empirical data are shown
        for MSL and GBR.
        (D--I) Marginal two-way spectra predicted by the model and their residuals 
        against the data are plotted for the three pairs of populations MSL, GBR 
        (D and G), MSL, Vindija (E and H), and GBR, Vindija (F, I).
    }
    \label{fig:humans}
\end{figure}

>>>>

\emph{Keywords:} Demographic inference, site frequency spectrum, diffusion
equation

@@

Many evolutionary mechanisms are known to affect allele frequency dynamics and
the SFS. This presents a challenge for demographic inference, as the observed
SFS may be distorted by non-demographic processes. However, this also presents
an opportunity to learn about different evolutionary processes, if demographic
history can be controlled for. The SFS has been used to infer the distribution
of fitness effects of new mutations, primarily for nonsynonymous variation in
coding regions (e.g., \cite{eyre2006distribution, boyko2008assessing,
kim2017inference}); to scan for historical selective sweeps (e.g.,
\cite{kim2002detecting, nielsen2005genomic}); and to understand how selection
on quantitative traits impacts the genetic architecture underlying those traits
(e.g., \cite{patel2024conditional, ragsdale2024archaic}). The SFS has also been
used to infer relative mutation rates and how they have changed over time
\cite{dewitt2021nonparametric}. Each of these analyses typically requires
partitioning the data in some way -- such as by functional annotation, mutation
context, local recombination rate -- so that comparisons can be made across SFS
from different classes of mutations.

@@

\singlespacing
\begin{minted}[gobble=4, frame=single,]{yaml}
    description: split-with-migration example model
    generation_time: 29
    time_units: years
    demes:
    - name: ancestral
      epochs:
      - {start_size: 15000, end_time: 300000}
      - {start_size: 25000, end_time: 75000}
    - name: popA
      ancestors: [ancestral]
      epochs:
      - {start_size: 30000}
    - name: popB
      ancestors: [ancestral]
      epochs:
      - {start_size: 1200, end_size: 14000}
    migrations:
    - demes: [popA, popB]
      rate: 5e-5
\end{minted}
\doublespacing

In \moments, a \demes-specified demographic model can be directly used to
compute expectations for either the SFS or multi-population LD statistics
\cite{ragsdale2019models, ragsdale2020unbiased}. For the SFS, we simply need
to load the demographic model using \demes and then specify the number of
haploid samples to draw and from which populations.
\begin{minted}[gobble=4, frame=single]{python}
    import demes, moments
    g = demes.load('model.yaml')
    samples = {'popA': 60, 'popB': 60}
    fs = moments.Demes.SFS(g, samples=samples)
\end{minted}
Here, we have not specified a mutation rate. The default behavior sets $u=1$,
so that $\theta = 4N_eu=4N_e$, where $N_e$ is the initial size of the ancestral
population. With a known mutation rate, the SFS will be properly scaled by
passing the mutation rate as a keyword argument: \texttt{moments.Demes.SFS(g,
samples=samples, u=u)}. This is equivalent to scaling the output SFS above by
multiplying by \texttt{u}, as \texttt{u*fs}.

@@

Below, we show two examples of multi-population demographic inference using the
joint SFS. The first example simulates data with known demographic parameters,
which we attempt to reinfer using both the original model and simpler
misspecified models. In the second example, we infer a relatively simple
Neanderthal-human history using data from two human populations and a
high-coverage ancient sample from the Neanderthal lineage. For both of these
examples, we briefly describe data processing and inference setup, emphasizing
the high-level components of the analyses. For detailed information about each
example, including managing data, specifying models, performing inference,
computing confidence intervals and visualizing fits, we refer readers to the
accompanying GitHub repository
(\url{https://github.com/StatisticalPopulationGenomics-2ndEd/moments}), which
is maintained with up-to-date versioning.

@@

We used publicly available data to infer a demographic model for two
modern human populations (a northern European and western African population) 
and a Neanderthal individual. As shown below, even with just a single sample 
from the Neanderthal lineage, we are able to recover key parameters such as the
Neanderthal-human divergence time and the fraction of ancestry contributed to 
modern Europeans by Neanderthals.

We obtained modern human genome sequences from a recent resequencing of the
1000 Genomes Project cohort \cite{byrska2022high}, and used the high-coverage
sequence of the Vindija Neanderthal genome \cite{prufer2017high}.  From the
1000 Genomes cohort we took 85 MSL (Mende from Sierre Leone) and 91 GBR
(British from England and Scotland) diploid sequences as our modern human
samples. We filtered out sites that lay outside the 1000 Genomes `strict' mask,
were not genotyped in four high-coverage archaic genome sequences, lacked
high-confidence ancestral state assignments, or fell within 10 kb of exonic or
promoter regions. This left us with approximately 960 Mb of well-characterized
and putatively neutrally-evolving sequence. We estimated the genome-wide SFS
using built-in \moments parsing functions. Details concerning data processing
(including our liftover of the Vindija sequence to a more recent genome build)
can be found in the linked GitHub repository.

@@

We next stitched these one-population models together and refitted the
resulting demographic model to get a two-population model that included
continuous symmetric migration between MSL and GBR and an ancestral size
expansion.  Following this, we added a Neanderthal branch and fitted the
Neanderthal-human divergence time, the Neanderthal effective size, and the
Neanderthal-GBR pulse proportion, leaving all other parameters fixed. We also
fitted the divergence time between the sampled Vindija Neanderthal lineage and
the Neanderthal lineage which admixed with non-African humans. Finding that
this parameter tended to converge to its lower bound (imposed by the estimated
time of death of the Vindija Neanderthal, ~55 kya), we fixed it to a value
supported by literature (90 kya, \cite{prufer2017high}) in a second round of
optimization. We also fixed the time of the Neanderthal-GBR pulse to 50 kya (at
the upper range of values inferrred by \cite{sumer2025earliest}).  Fixing
parameters for which good estimates are available in the literature is often
reasonable in the first stages of model optimization, and we expected this
value to be especially poorly constrained due to the small pulse proportion and
Neanderthal sample size. In the second round of optimization of the
three-population model, we refitted all parameters aside from the two times
mentioned above to refine the estimates.

After this refitting step, we were left with parameters in good agreement with
those inferred by prior authors, with a Neanderthal-human divergence time of
547 kya, MSL-GBR split time of 77 kya, and Neanderthal-GBR pulse proportion of
2.7 percent. We observe some large, systematic residuals in our best-fit model
(Figure~\ref{fig:humans}G-I), suggesting model misspecification that could be
improved through further exploration of model space.

We also fitted a nested model without the Neanderthal-GBR pulse, which had a
substantially lower likelihood than the pulse model (log-likelihood [LL]
$\approx -127,552$, against LL $\approx -25,103$ for the pulse model). We did
not perform a formal hypothesis test; the null distribution of the
likelihood-ratio test (LRT) statistic is overdispersed under a composite
likelihood, inflating the rate of false positives and preventing its use in
this setting. \moments does incorporate a function
(\texttt{moments.Godambe.LRT\_adjust()}) which computes a multiplicative
adjustment to the LRT, intended to restore the expected null distribution,
using first-order moment matching \cite{coffman2016computationally}. This
method requires bootstrapping over the data. Otherwise, we could conduct a
hypothesis test by fitting the nested and pulse models to a large number of
bootstrap samples and calculating the rate of rejection across them. This has
the benefit of producing a distribution of (nonindependent) LRT statistics, but
is time-consuming and computationally expensive.

@@

\begin{figure}
    %\includegraphics{../example1/fig1.pdf}
    \caption{
        \textbf{Inference using simulated data.}
        (A) The true demographic model, including an expansion in the ancestral
        population and symmetric migration between the descendent populations.
        (B--D) The simulated data, showing both marginal and the joint SFS.
        (E--G) Residuals between the data SFS (in D) and the predicted SFS from
        the inferred demographic models in (H--J), resp. The two misspecified
        models (H and I) provide worse fits to the data and both overestimate
        the split time due to disallowing any size change in the ancestral
        population.
    }
    \label{fig:im}
\end{figure}

\begin{figure}
    %\includegraphics{../example2/fig2.pdf}
    \caption{
        \textbf{Inference using human and Neanderthal data.}
        (A) The best-fit model including the Vindija Neanderthal and two human
        populations.
        (B, C) Marginal SFS fits between the model and empirical data are shown
        for MSL and GBR.
        (D--I) Marginal two-way spectra predicted by the model and their residuals 
        against the data are plotted for the three pairs of populations MSL, GBR 
        (D and G), MSL, Vindija (E and H), and GBR, Vindija (F, I).
    }
    \label{fig:humans}
\end{figure}
